Reproducable Research Week 2 - CP1
=====================

## Loading libraries 
(sorry for the output)

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(gridExtra)
```

## Loading and preprocessing the data

```{r echo=TRUE}
activity <- read.csv("./data/activity.csv", stringsAsFactors =FALSE)
activity$date <- as.Date(as.character((activity$date )), "%Y-%m-%d")
str(activity)
```

## What is mean total number of steps taken per day?

### Calculate the total number of steps taken per day

```{r echo=TRUE}
days <- group_by(activity, date)
activity.perday <- summarize(days, steps = sum(steps, na.rm = TRUE) )
head(activity.perday)
```

### Make a histogram of the total number of steps taken each day

```{r echo=TRUE}
qplot(activity.perday$steps, geom="histogram",  main = "Histogram for Total steps per day", xlab = "Steps", ylab = "Days Count", bins = 30) 
```

### Calculate and report the mean and median of the total number of steps taken per day

Mean of total steps taken per day

```{r echo=TRUE}
mean(activity.perday$steps)
```

Median of total steps taken per day

```{r echo=TRUE}
median(activity.perday$steps)
```

## What is the average daily activity pattern?

Time series plot of the 5 minunte interval and average number of steps taken, averaged across all days:

```{r echo=TRUE}
intervals <- group_by(activity, interval)
activity.perinterval <- summarize(intervals, steps = mean(steps, na.rm = TRUE) )
##ggplot(activity.perinterval, aes(interval, steps)) + geom_line() + xlab("interval (time)") + ylab("average steps")
activity.perinterval <- mutate(activity.perinterval, time = hms(paste(as.character(interval %/% 100), as.character(interval %% 100), "00" , sep=":")))
qplot( as.numeric(time)/300, steps, data = activity.perinterval , geom="line", xlab = 'intervals (5 min intervals)')
## qplot(interval, steps, data = activity.perinterval , geom="line")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}
arrange(activity.perinterval, desc(steps))[1,1:2]
```

We see that interval 835 has the most steps on average across all days

## Imputing missing values

How many missing days/intervals are missing (coded as NA) values?

```{r echo=TRUE}
sum(is.na(activity$steps))
```

We will fill the missing values by taking the mean of that interval:

```{r echo=TRUE}
activity.imputed <- activity

activity.imputed <- merge(activity.imputed, activity.perinterval, by = "interval")
name <- "time"
activity.imputed <- activity.imputed %>% select(-one_of(name))
activity.imputed <- rename(activity.imputed, steps = steps.x)
activity.imputed$steps[is.na(activity.imputed$steps)] <- as.integer(activity.imputed$steps.y[is.na(activity.imputed$steps)])
name <- "steps.y"
activity.imputed <- activity.imputed %>% select(-one_of(name))
activity.imputed <- arrange(activity.imputed,date)
head(activity.imputed)
```

A histrogram of the total number of steps taken each day

```{r echo=TRUE}
days.imputed <- group_by(activity.imputed, date)
activity.imputed.perday <- summarize(days.imputed, steps = sum(steps, na.rm = TRUE) )
qplot(activity.imputed.perday$steps, geom="histogram",  main = "Histogram for Total steps per day", xlab = "Steps", ylab = "Days Count", bins = 30) 
```

Mean total number of steps taken per day

```{r echo=TRUE}
mean(activity.imputed.perday$steps)
```

Median total number of steps taken per day

```{r echo=TRUE}
median(activity.imputed.perday$steps)
```

We see that imputing missing data slightly increased the estimates of total number of steps taken.

## Are there differences in activity patterns between weekdays and weekends?

We will create a factor variable 'day.type' in the dataset indicating if it's a weekday or weekend

```{r echo=TRUE}
activity.imputed = mutate(activity.imputed, day.name = weekdays(activity.imputed$date), day.type = as.factor(ifelse(day.name %in% c("zaterdag", "zondag"), "weekend", "weekday")))
name <- "day.name"
activity.imputed <- activity.imputed %>% select(-one_of(name))
head(activity.imputed)
```

Now we will make a plot comparing activity between weedays and weekend

```{r echo=TRUE}
activity.weekend <- subset(activity.imputed, ! day.type %in% c("weekday"))
intervals.weekend <- group_by(activity.weekend, interval)
activity.weekend.perinterval <- summarize(intervals.weekend, steps = mean(steps, na.rm = TRUE) )
activity.weekend.perinterval <- mutate(activity.weekend.perinterval, time = hms(paste(as.character(interval %/% 100), as.character(interval %% 100), "00" , sep=":")))

activity.weekday <- subset(activity.imputed, ! day.type %in% c("weekend"))
intervals.weekday <- group_by(activity.weekday, interval)
activity.weekday.perinterval <- summarize(intervals.weekday, steps = mean(steps, na.rm = TRUE) )
activity.weekday.perinterval <- mutate(activity.weekday.perinterval, time = hms(paste(as.character(interval %/% 100), as.character(interval %% 100), "00" , sep=":")))

plot1 <- qplot( as.numeric(time)/300, steps, data = activity.weekend.perinterval , geom="line", xlab = 'intervals (5 min intervals)', main = "Weekend")
plot2 <- qplot( as.numeric(time)/300, steps, data = activity.weekday.perinterval , geom="line", xlab = 'intervals (5 min intervals)', main = "Weekday")
grid.arrange(plot1, plot2, nrow=2, ncol=1)
```

It seems that that activity start earlier on weekdays (prepping for work?) than in the weekend. During the rest of the day there's more activity in the weekend.

